---
title: "Tools of the Trade"
subtitle: "HES 505 Fall 2025: Session 2"
author: "Matt Williamson"
format: 
  revealjs:
    theme: mytheme.scss
    highlight-style: pygments
    slide-number: true
    show-slide-number: print
    embed-resources: true 
---
```{r}
#| echo: false
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
library(tidyverse)
```
## Checking in

1. What can I clarify about the course?


## Today's Plan {background-image="img/slide_2/hudbay.jpeg" background-opacity="0.4"}

:::{style="font-size: 1.4em; text-align: middle; margin-top: 2em"}

- What do we mean by reproducibility?

- What is version control and why is it helpful?

- What is a (spatial) data workflow?

- Why (not) R?

:::

# Why (not) R? {background="#9F281A"}

## Why R?

::: columns
::: {.column width="30%"}
::: incremental

-   Open Source

-   Huge useR community

-   Integrated analysis pipelines

-   Reproducible workflows
:::
:::

::: {.column width="70%"}

::: panel-tabset
### Code

```{r}
#| echo: true
library(maps)
library(socviz)
library(tidyverse)
party_colors <- c("#2E74C0", "#CB454A") 
us_states <- map_data("state")
election$region <- tolower(election$state)
us_states_elec <- left_join(us_states, election)
p0 <- ggplot(data = us_states_elec,
             mapping = aes(x = long, y = lat,
                           group = group, 
                           fill = party))
p1 <- p0 + geom_polygon(color = "gray90", 
                        size = 0.1) +
    coord_map(projection = "albers", 
              lat0 = 39, lat1 = 45) 
p2 <- p1 + scale_fill_manual(values = party_colors) +
    labs(title = "Election Results 2016", 
         fill = NULL)
```

### Plot

```{r}
#| echo: false
p2
```
:::
:::
:::

## Why not R? 

```{r}
#| code-overflow: wrap
#| echo: true

## ---
## Error: could not find function "performance"
## ---
##  [1] "Error in if (str_count(string = f[[j]], 
##  pattern = \"\\\\S+\") == 1) 
##  { : \n  argument is of length zero"   
## ---
## Error in eval(expr, envir, enclos) : object 'x' not found
## ---
## Error in file(file, "rt") : cannot open the connection
## ---

```

::: columns
::: {.column width="50%"}
-   Coding can be hard...

-   Memory challenges
:::
::: {.column width="50%"}
-   Speed

-   Decision fatigue
:::
:::

## Getting Help
::: columns
::: {.column width="50%"}
- Google it!! 
  - Use the exact error message
  - Include the package name
  - include "R" in the search

:::
::: {.column width="50%"}
- Stack Overflow
  - Reproducible examples 
- Package "issue" pages
- r_spatial slack channel
- [Common errors](https://github.com/noamross/zero-dependency-problems/blob/master/misc/stack-overflow-common-r-errors.md)

:::
:::

::: {style="font-size: 1.2em; text-align: center; color: #9F281A"} 
**Ask Me**
:::




