---
title: "Probabilistic Interpolation"
subtitle: "HES 505 Fall 2025: Session 20"
author: "Matt Williamson"
format: 
  revealjs:
    theme: mytheme.scss
    highlight-style: pygments
    slide-number: true
    show-slide-number: print
    embed-resources: true  
---

```{r}
#| include: false
library(terra)
library(sf)
library(tidyverse)
library(gstat)  # Used for the dirichlet tesselation function
```

## Interpolation

>Spatial interpolation is the process of predicting values of a target variable or attribute at an unobserved location, from a set of values at known (observed) locations

## Deterministic vs. Probabalistic Interpolation

::: incremental

* **Deterministic**- interpolated values calculated as a function of distance

* **Probabilistic**- interpolated values are _predicted_ based on statistical models
:::

## Kriging

::: {style="font-size: 0.8em"}
* Previous methods predict $z$ as a (weighted) function of distance

* Treat the observations as perfect (no error)

* If we imagine that $z$ is the outcome of some spatial process such that:

$$
\begin{equation}
z(\mathbf{x}) = \mu(\mathbf{x}) + \epsilon(\mathbf{x})
\end{equation}
$$

then any observed value of $z$ is some function of the process ($\mu(\mathbf{x})$) and some error ($\epsilon(\mathbf{x})$)

* Kriging exploits autocorrelation in $\epsilon(\mathbf{x})$ to identify the trend and interpolate accordingly

:::

## Autocorrelation

* __Correlation__ the tendency for two variables to be related

* __Autocorrelation__ the tendency for observations that are closer (in space or time) to be correlated

* __Positive autocorrelation__ neighboring observations have $\epsilon$ with the same sign

* __Negative autocorrelation__ neighboring observations have $\epsilon$ with a different sign (rare in geography)

## Ordinary Kriging

* Assumes that the deterministic part of the process ($\mu(\mathbf{x})$) is an unknown constant ($\mu$)

$$
\begin{equation}
z(\mathbf{x}) = \mu + \epsilon(\mathbf{x})
\end{equation}
$$

## Steps for Ordinary Kriging

* Removing any **spatial trend** in the data (if present).
** Computing the **experimental variogram**, $\gamma$, which is a measure of spatial autocorrelation.
* Defining an **experimental variogram model** that best characterizes the spatial autocorrelation in the data.
* Interpolating the surface using the experimental variogram.

## (semi)Variograms

* Relationship between observation variance and distance

* __nugget__ - the proportion of semivariance that occurs at small distances

* __sill__ - the maximum semivariance between pairs of observations

* __range__ - the distance at which the __sill__ occurs 

## Variograms

* Empirical vs. fitted.

```{r}
#| echo: false

show.vgms(par.strip.text = list(cex = 0.75))

```


## Universal Kriging

* Assumes that the deterministic part of the process ($\mu(\mathbf{x})$) is now a function of the location $\mathbf{x}$

* Could be the location or some other attribute

* Now `y` is a function of some aspect of `x`

## Co-Kriging

* relies on autocorrelation in $\epsilon_1(\mathbf{x})$ for $z_1$ AND cross correlation with other variables ($z_{2...j}$)

* Extending the ordinary kriging model gives:

$$
\begin{equation}
z_1(\mathbf{x}) = \mu_1 + \epsilon_1(\mathbf{x})\\
z_2(\mathbf{x}) = \mu_2 + \epsilon_2(\mathbf{x})
\end{equation}
$$

* Note that there is autocorrelation within both $z_1$ and $z_2$ (because of the $\epsilon$) and cross-correlation (because of the location, $\mathbf{x}$)

* Not required that all variables are measured at exactly the same points

## Why Krig?

* Allows empirical estimate of autocorrelation

* Incorporate actual spatial structure

* Increased accuracy

* Uncertainty estimates

## Why Not Krig?

* Computationally intensive

* Smoothing not guarenteed

* Variograms (somewhat) subjective

