---
title: "Meeting Your Data"
date: "2025-09-03"
---
```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

Now that you know a bit about the "why" and "how" of this course, it's time to start actually working with data. We'll spend a bit of time thinking about the 'nature' of data, the qualities of "good" data, and the responsibilities of data creators and users. We'll also practice generic approaches for getting your own data into `R` and accessing data via APIs and functions.

## Readings

{{< fa arrow-up-right-from-square >}} [Chapter 2](https://r.geocompx.org/spatial-class) in Geocomputation with R [@lovelace_geocomputation_2019] provides and overview of using `sf` for vector datasets and `terra` for raster data.

{{< fa arrow-up-right-from-square >}} [Chapter 2](https://www.paulamoraga.com/book-spatial/spatial-data-in-r.html) from Moraga (2023) explores similar topics, but provides more explanation about projections, coordinates, etc.

{{< fa arrow-up-right-from-square >}} [Simple Features for R](https://r-spatial.github.io/sf/articles/sf1.html) provides a more in-depth, technical discussion of how the `sf` package organizes spatial data and attributes.

{{< fa arrow-up-right-from-square >}} [Ch 2.1-2.5 from Spatial Data Science with R and terra](https://rspatial.org/spatial/index.html) describes the basic functionality of `terra` for both vector and raster datasets. For reasons we'll discuss in class, we will rarely use `terra` for vector data.

## Objectives

By the end of today, you should be able to:

1. **Summarize** key debates surrounding the role of (spatial) data in solving environmental problems 

2. **Describe** FAIR and CARE principles for data and their relationship to existing debates

3. **Read** data into your `R` environment.

4. **Inspect** the data and summarize it using tables and simple plots


## Slides
The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/04-slides")
```
<div>

```{=html}
<iframe class="slide-deck" src="/slides/04-slides.html"></iframe>
```

</div>