---
title: "Data Manipulation with the `tidyverse`"
date: "2025-09-08"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

We've established some of the reasons why data reproducibility and transparency are important. We've even gotten to the point where you can bring data into `R` from a variety of sources. Now we've actually got to start wrangling it into the sorts of things we can use for analysis. For a variety of reasons, this course relies on the `tidyverse` family of packages and its extensions into the spatial realm. Today, we'll talk about why and begin to develop some intuition for building an analysis using the functions contained in the `tidyverse` family of packages.

## Readings

{{< fa arrow-up-right-from-square >}} 

{{< fa arrow-up-right-from-square >}} 

{{< fa arrow-up-right-from-square >}} 

## Objectives
By the end of today you should be able to:


* **Explain** the importance of readable code.

* **Articulate** the structure of readable scripts

* **Utilize** the verb, object, helper syntax of the `tidyverse` to modify your data in a reproducible way


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/03-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/03-slides.html"></iframe>
```

</div>