---
title: "Multivariate Description"
date: "2025-10-22"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

Now that we've looked at describing spatial patterns using point processes and patch metrics, we'll finish by taking a look at multivariate methods for making sense of spatial patterns. We'll focus on two, relatively simple approaches. Principal components analysis for identifying and mapping latent gradients in quantitative data and clustering for identifying qualitative 'groups' in data based on quantitative data. These are just two approaches in a broader class of "dimension reduction" methods, but they are illustrative of the basic tasks of multivariate description.

## Resources

### Bigger Picture

- <i class="fas fa-external-link-square-alt"></i> [Principal Component Analysis on Spatial Data: An Overview](https://www.tandfonline.com/doi/full/10.1080/00045608.2012.689236#d1e979) by [@demvsar2013] provides a semi-technical introduction to why we might use Principle Components Analysis to identify spatial gradients and things we should consider when applying PCA to spatial data.

- <i class="fas fa-external-link-square-alt"></i> [A latitudinal gradient in dimensionality of biodiversity](https://nsojournals.onlinelibrary.wiley.com/doi/epdf/10.1111/ecog.03654) by [@stevens2018] is a nice, contemporary demonstration of PCA for ecological questions and spatial data.

- <i class="fas fa-external-link-square-alt"></i> [Archetypical patterns and trajectories of land systems in Europe](https://link.springer.com/article/10.1007/s10113-015-0907-x) by [@levers2018] is a great reference for using cluster analysis to identify social-ecological archetypes for sustainability problems.

- <i class="fas fa-external-link-square-alt"></i> [The American West as a social-ecological region: drivers, dynamics and implications for nested social-ecological systems](https://iopscience.iop.org/article/10.1088/1748-9326/ab4562) by [@jones2019american] is an example of a fairly simple cluster analysis that resulted from a workshop hosted here at Boise State!

### Technical Details

- <i class="fas fa-external-link-square-alt"></i> [Cluster Analysis](https://www.erikkusch.com/courses/bftp-biome-detection/cluster-analysis/) provides a nice vignette introducing the workflow for cluster analysis. It's done using the `raster` package (a precursor to `terra`), but the translation into `terra` is straightforward.

- <i class="fas fa-external-link-square-alt"></i> [Principal Components Analysis](https://rspatial.org/rs/3-basicmath.html#principal-component-analysis) from the online `terra` workbook provides an intro to the workflow for a simple PCA using rasters.

## Objectives

By the end of today you should be able to:

- **Describe** the motivations for dimension reduction with spatial data

- **Distinguish** the differences between PCA and cluster analysis

- **Understand** common metrics for assessing the quality of dimension reduction results 

- **Implement** simple PCA and cluster analysis using `R` 


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/18-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/18-slides.html"></iframe>
```

</div>

