---
title: "Interpolation"
date: "2025-10-27"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

There are many cases in the analysis of social and ecological data where we might be missing data from a population or region where we hope to make inference or conduct further analysis. When the data is not spatial, we use _imputation_ as a statistically principled approach for filling in these gaps. When the data is spatial, we use **interpolation** as a way of assigning values to locations where the data is missing such that we leverage the fact that most data is spatially autocorrelated. There are two general approaches for assigning these values: _deterministic_ and _probabilistic_. Today we'll take a look at deterministic approaches for interpolating missing data.


## Resources

### Bigger Picture


- <i class="fas fa-external-link-square-alt"></i>  [Can niche-based distribution models outperform spatial interpolation?](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1466-8238.2007.00331.x) by [@bahn2007] illustrates the power of spatial interpolation for species distribution modeling and the potential for spurious inferences when failing to account for spatial structure in species data.

- <i class="fas fa-external-link-square-alt"></i> [Social–ecological hotspots mapping: A spatial approach for identifying coupled social–ecological space](https://www.sciencedirect.com/science/article/pii/S0169204607002216) by [@alessa2008] uses interpolation to map human values across a landscape and provides an interesting justificaiton for why we might prefer this approach over something more statistically sophisticated.

### Technical Details

- <i class="fas fa-external-link-square-alt"></i> The [Spatial Interpolation](https://r-spatial.org/book/12-Interpolation.html) chapter of [@sds] provides some code for using `sf` and `stars` to create spatial interpolation. We'll be replacing the raster operations with `terra` functions, but the workflow is the same.

- <i class="fas fa-external-link-square-alt"></i> [Spatial interpolation in R](https://mgimond.github.io/Spatial/interpolation-in-r.html) from Manuel Gimond provides a less statistical overview of the same workflow and uses `terra` instead of `stars`.

## Objectives

By the end of today you should be able to:

*   **Describe** reasons for interpolating data

*   **Distinguish** between deterministic and probabilistic methods for interpolation

*   **Characterize** the difference between distance- and spline-based interpolation

*   **Implement** two common interpolation methods in `R`


```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/19-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/19-slides.html"></iframe>
```

</div>

