---
title: "Probabilistic Interpolation"
date: "2025-10-29"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))


```



## Resources

### Technical Details

- <i class="fas fa-external-link-square-alt"></i> [Ch. 14: Kriging](https://www.paulamoraga.com/book-spatial/kriging.html) in
from Paula Moraga's new book Spatial Statistics for Data Science: Theory and Practice with `R` gives a little gentler introduction to spatial neighbors specifically in the context of statistical models.

- <i class="fas fa-book"></i> [Chapter 12: Spatial Interpolation](https://r-spatial.org/book/12-Interpolation.html) in Spatial Data Science by Edzer Pebesma and Roger Bivand provides some additional details for managing `gstat` and `sf`, but relies on `stars` a package we don't use in this course.

## Objectives

By the end of today you should be able to:

* Describe and implement statistical approaches to interpolation

* Describe the case for identifying neighbors with areal data

* Implement contiguity-based neighborhood detection approaches

* Implement graph-based neighborhood detection approaches



```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/20-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/20-slides.html"></iframe>
```

</div>

