---
title: "Statistical Modelling I"
date: "2025-11-10"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

Last class we looked at tree-based machine-learning models for classifying data (i.e., presence/abscence, categorization) or estimating the pseudo-regression relationships between variables and measured outcomes. Those approaches rely on learning the distributions necessary for the likelihood from the data themselves (rather than from a theoretical probability distribution). Today we'll look at alternative approaches using tradititional statistical models.

## Resources

### Bigger Picture

- <i class="fas fa-external-link-square-alt"></i> [Logistic regression: a brief primer](https://onlinelibrary.wiley.com/doi/10.1111/j.1553-2712.2011.01185.x) by [@Stoltzfus2011] is a nice introduction to logistic regression.

- <i class="fas fa-external-link-square-alt"></i> [Is my species distribution model fit for purpose? Matching data and models to applications](https://onlinelibrary.wiley.com/doi/full/10.1111/geb.12268) by [@Guillera-Arroita2015] is an excellent, concise description of the relations between data collection, statistical models, and inference.

- <i class="fas fa-external-link-square-alt"></i> [Predicting species distributions for conservation decisions](https://onlinelibrary.wiley.com/doi/full/10.1111/ele.12189) by [@Guisan2013] is a foundational paper describing some of the challenges with making conservation decisions based on the outcomes of species distribution models.

### Techincal details

- <i class="fas fa-external-link-square-alt"></i> [Statistical Models](https://rspatial.org/intr/13-statmodels.html) from the `terra` package documentation has some step-by-step examples of using `terra` with base `R` to fit statistical modles.

- <i class="fas fa-external-link-square-alt"></i> [Statistical Learning](https://r.geocompx.org/spatial-cv) from [@lovelace_geocomputation_2019] has an extended example of fitting a variety of modeling approaches to data and evaluating their performance.


## Objectives

By the end of today you should be able to:

- **Define** the likelihood function and its relationship to statistical infrerence.

- **Recognize** key assumptions of statistical models and how spatial data may challenge those assumptions.

- **Simulate** fake data with known relationships

- **Fit** simple linear and generalized linear models to spatial data

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/23-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/23-slides.html"></iframe>
```

</div>

