---
title: "Data Models, Coordinates, and Geometries"
---

## Objectives

We're going to introduce the `sf`, `terra`, and `tidyterra` packages today as a way of reading and inspecting spatial data. By the end of this lesson, you should be able to:

1. Read in spatial data from both the vector and raster data model.

2. Recognize the required components of a spatial data object

3. Convert a spreadsheet into spatial data

4. Use `plot` as a way to perform a quick check on your data.



## Getting Started

First, [join the repository](https://classroom.github.com/a/X4ivBDX3). Then, we'll have to [get our credentials](https://rstudio.aws.boisestate.edu/) and login to a [new RStudio session](https://rstudio.apps.aws.boisestate.edu/). Finally, create a new, version control, git-based project (using the link to the repo you just created). 

Once you've done that, you'll need to load your packages. We're going to add 3 new packages to the `tidyverse`; the `sf`, `terra`, and `tidyterra` packages. We'll also add the `googledrive` so that we can use a custom function for downloading data from a shared Google Drive folder. Note that when you load these, the conflicts that the `tidyverse` reports will be different. In order to overcome these conflicts, you'll need to use explicit calls like `package::function` so `R` knows which ones to use.

```{r ldpkg}
library(googledrive)
library(sf)
library(terra)
library(tidyterra)
library(tidyverse)
```

## Loading a Custom Function

You'll notice that your `scripts/utilities` folder has a file called `download_utils.R`. This file has some custom functions for downloading data directly from websites and from a Google Drive Folder. In order to give yourself access to these functions you'll need to `source` them into your environment. You can do this by typing:

```{r srcfun}
source("scripts/utilities/download_utils.R")

```

The first time you run the `gdrive_folder_download` function, you'll be taken to a webpage where you need to give `googledrive` access to your Google Drive folders. Make sure you've checked all the boxes and click Allow. You should get a message saying that authorization is complete and that you can close the tab.


## Reading Data: Vectors
