---
title: "Projections, Extent, and Resolution"
---

## Objectives

Now that you know how to read in some spatial data, we'll spend a bit of time learning the `sf` and `terra` syntax for accessing and changing the projection, extent, and resolution. By the end of this lesson, you should be able to:

1. Access the current CRS, extent, and resolution of a spatial data object

2. Use `sf` and `terra` to modify 1 spatial attribute or all 3 at once

3. Align vector and raster data for analysis and combination



## Getting Started

First, [join the repository](https://classroom.github.com/a/XWzfBM16). Then, we'll have to [get our credentials](https://rstudio.aws.boisestate.edu/) and login to a [new RStudio session](https://rstudio.apps.aws.boisestate.edu/). Finally, create a new, version control, git-based project (using the link to the repo you just created). 

Once you've done that, you'll need to load your packages. We're going to use the same packages as last time and source our custom download function.

```{r ldpkg}
library(googledrive)
library(sf)
library(terra)
library(tidyterra)
library(tidyverse)
source("scripts/utilities/download_utils.R")
```

## Some Context

You likely noticed in your quick maps that the two datasets look different. Sure, one is a vector with lots of `POLYGON` elements and the other is a raster (stored as a `SpatRaster`). If you looked a little, you probably noticed that the orienation (angle) of the areas was a bit different. If you looked a little harder, you probably saw that the general shapes of the states is a bit different, too. And, maybe if you squint, you can see that the boundaries of the maps looks a little different. We'll work through some ways of looking at and fixing those things today! Let's load our datasets and get started!

```{r dlld}
file_list <- drive_ls(drive_get(as_id("https://drive.google.com/drive/u/1/folders/1PwfUX2hnJlbnqBe3qvl33tFBImtNFTuR")))


file_list %>% 
  dplyr::select(id, name) %>%
  purrr::pmap(function(id, name) {
    gdrive_folder_download(list(id = id, name = name),
                           dstfldr = "inclass08")
  })

cejst_shp <- read_sf("data/original/inclass08/cejst.shp")
fire_prob <- rast("data/original/inclass08/wildfire.tif")
```


## What is the Current Extent?

Because extent is part of our internal definition of scale and because getting our extents aligned is a big part of making sure our data sandwich doesn't fall apart, we'll start by checking to see what the extent is for each dataset.

### Vector Data
By convention, the `sf` package uses a standard syntax for spatial operations that starts with `st_`. In addition, the extent of a vector dataset is known as the **bounding box**. We can access that using `st_bbox`



```{r sfbbox}

st_bbox(cejst_shp)

cejst_shp %>% 
  filter(., SF == "Idaho") %>% 
  st_bbox()

```
There are three things to know about the bounding box. First, it is, by definition, a rectangle that encompases the vertices of your vector object. So, even if you've got an irregularly shaped `POLYGON` or a series of `POINTS`, the bounding box  necessarily returns the rectangle that encompasses all of the vertices in the object.  Second, using `st_bbox` returns an object of class `bbox` (you can check this by running `st_bbox(your_data) %>% class()` which is a special kind of object. Finally, if you use `dplyr` verbs to alter your the composition of your vector dataset, the bounding box will update to reflect those changes. You can see that in the values returned in the example above, but you can also plot it.

To do that, we have to convert the `bbox`to a simple feature collection (`sfc`) we can do that using the `st_as_sfc` command and then use base `plot` to give us a sense for how different the two bounding boxes are.

```{r pltbbox}
all_bbox <- st_bbox(cejst_shp) %>% 
  st_as_sfc()
id_bbox <- cejst_shp %>% 
  filter(., SF == "Idaho") %>% 
  st_bbox() %>% 
  st_as_sfc()

plot(all_bbox)
plot(id_bbox, add=TRUE, border ="red")
```


One nice thing about `sf` objects is that there is a built-in `geom_` in `ggplot2` that will allow us to make maps with all of the flexibility of `ggplot`. 

```{r ggbbox}

ggplot() +
  geom_sf(data = all_bbox, color = "yellow", fill = "darkorchid4") +
  geom_sf(data= id_bbox, fill = "orangered") +
  theme_bw()

```
In the `ggplot` call, I've done several things. First, I used `geom_sf` to add each vector dataset. Then instead of using the `mapping = aes()` to tell `ggplot` which column I want to use for the colors, I set the `color` and `fill` manually by specifying them outside of the `aes()` call and setting them manually.  Finally, I called `theme_bw()` to set the plotting options [using one of the preset `ggplot` themes](https://r-charts.com/ggplot2/themes/). Using `ggplot` gives a lot more flexibility and allows us to layer data on top of each other to create maps!


### Raster Data

Unfortunately, the `terra` package uses some different conventions and `tidyterra` only provides functionality for some operations that are more complex. That said, many of the functions in `terra` match our intuition for the operations we'd like to use so it doesn't take too much to figure out how to get what we need. In this case, we are interested in the **extent** of our raster.  If you look at the `terra` helpfile (`?terra`), you'll see the entire list of functions available within the package. If you look for the "Getting and setting dimensions" section, you'll see that there are a number of functions for getting information about `SpatRaster` objects. A quick scan reveals that the function for getting the extent is `ext`.

```{r terext}
ext(fire_prob)

```
You'll notice a few things about this object. First, it has a new `class`, a `SpatExtent`. Next, you'll notice that the ordering and structure of the data is different from that of `sf`'s `bbox`. 

We can use `plot` directly on a `SpatExtent` object to be able to look at the extent.

```{r plotext}
plot(ext(fire_prob), border = "red")
plot(fire_prob, add=TRUE)
```

Similar to the `bbox` from `sf`, `ext` returns a rectangle encompassing the entire `SpatRaster` object. The `tidyterra` package allows us to use `filter` as a way to subset rasters without using indexing. We can use that here to find all of the cells whose wildfire hazard potential is greater than 30000. While the behavior of `filter` is the same for both `sf` and `terra` (meaning it keeps only the values that meet your `filter` criteria), you'll notice that, unlike the `bbox` of `sf`, the extent of a `SpatRaster` does not change to match the subset data. 

```{r subrst}

fire_subset <- fire_prob %>% 
  filter(WHP_ID > 30000)

plot(ext(fire_subset), border = "red")
plot(fire_subset, add=TRUE)

```

## How Can I Change the Extent?

There may be occasions where you want to manually change the extent of your spatial objects. Both `sf` and `terra` provide multiple functions for changing the `extent` as part of additional spatial operations (we'll talk more about this in the coming weeks), but sometimes you might want to adjust the extent of an object without modifying anything else about it. Sometimes we might do this to create inset maps. Sometimes we might want to create asymmetric buffers around an existing extent. This is a somewhat rare, but it's worth knowing how to do it just to get a feel for how extents work.

### Vector Data

If you look at `?st_bbox` you'll see an example of the manual creation of a `bbox` object. We can exploit the fact that `st_bbox` returns a `named.vector` to access the values in the original bounding box and "shrink" it by adding (or subtracting) values from the existing bounding box. Two things are tricky here. First, the original data is in a **geodetic** projection so the units are in degrees (which I have no real intuition for). Second, the goal of "shrinking" the bounding box means we have to think about how we might move "up" or "in" along the graticules. I exploited the plot above to figure out some reasonable numbers and add them here. 

```{r chgbbox}
orig_bbox <- st_bbox(cejst_shp)

new_bbox <- st_bbox(c(orig_bbox["xmin"] + 1, orig_bbox["xmax"] - 1,  orig_bbox["ymax"] - 1, orig_bbox["ymin"] + 1), crs = st_crs(orig_bbox)) %>% st_as_sfc()
```

You can verify this again with a plot.

```{r bboxchg}

ggplot() +
  geom_sf(data = st_as_sfc(orig_bbox), color = "yellow", fill = "darkorchid4") +
  geom_sf(data= new_bbox, fill = "orangered") +
  theme_bw()
  
```  
### Raster Data
The process is similar for rasters with `terra`
## What is the Current Resolution?
### Vector Data
### Raster Data
## How Can I Change the Current Resolution?
### Vector Data
### Raster Data
## What is the Current Projection?
### Vector Data
### Raster Data
## How Can I Change the Projection?
### Vector Data
### Raster Data

## A Note on Assigning Spatial Attributes