---
title: "Using Spatial Operations to Change Geometries and Explore MAUP"
---

## Objectives

1. Practice spatial operations to modify vector and raster data

2. Practice simulating data to explore ideas

3. Explore the Modifiable Areal Unit Problem with known data.

## Getting Started

We'll start like we usually do, by loading packages. There are a few new packages here. The `mvtnorm` packages is a way for generating random numbers from a multivariate normal distribution (like `rnorm`). The `patchwork` package provides a lot of functionality for making multipanel plots. We'll also `source` our utility function.

```{r}
library(sf)
library(terra)
library(mvtnorm)
library(tidyverse)
library(patchwork)
library(googledrive)
source("scripts/utilities/download_utils.R")

```

```{r dldata}

file_list <- drive_ls(drive_get(as_id("https://drive.google.com/drive/u/1/folders/1PwfUX2hnJlbnqBe3qvl33tFBImtNFTuR")))


file_list %>% 
  dplyr::select(id, name) %>%
  purrr::pmap(function(id, name) {
    gdrive_folder_download(list(id = id, name = name),
                           dstfldr = "inclass10")
  })
```

## Loading the Data We Have

There are two different vector datasets we're going to use for this example. The first (`pa.shp`) is our old friend, the Purple Air sensors. This is a `POINT` dataset. The second (`blockgroupPop.shp`) contains a number of `POLYGONS` for the Census block groups in Ada, Gem, Canyon, Elmore, and Owyhee counties in Idaho along with an attribute for their total population as of the 2020 Census. We'll load those here and take a quick look.

```{r lddata}

population_data <- read_sf("data/original/inclass10/blockgroupPop.shp")

pa_locs <- read_sf("data/original/inclass10/pa.shp") %>% 
  st_transform(., crs=crs(population_data)) %>% 
  st_crop(., population_data)

glimpse(population_data)
plot(population_data$geometry)
glimpse(pa_locs)
plot(pa_locs$geometry)

```


## Simulating Some Data

In order to explore how MAUP plays out in an analysis, we need some data where the relationship between the two datasets **is known**. Simulation is a great way to gain some intuition for how things work without getting trapped by the subtleties and imperfections of a dataset. Plus, if you're ever fitting complicated statistical models, running them first on data where you know the relationships (because you made them) is a great way to make sure your model is well specified.

There are a few things happening here, so I'll walk you through it.
First, we set a "seed" (`set.seed()`) - this ensures that the random numbers you generate on your computer are the same as those generated on my computer. It also ensures that no matter how many times you run this, you should get the same outputs. This is great for reproducibility, but not necessarily for sensitivity (where we want to make sure that our results aren't just a function of the one set of random numbers we drew).

Next we have to create a raster template that we can fill in with the random data we generate. You'll notice that we used `rast` (the same function we use to read in existing raster data) to do this **and** that we use the CRS (`crs()`) and extent (`ext()`) of our `population_data` shapefile. To make sure all of this works in the `terra` structure, we convert (temporarily) our `population_data` object to a `SpatVector` (using `vect()`). Finally, we use `ncell()` to count the total number of cells in our template raster so that we can make sure we draw enough values.

Now we want to create some data that is random, but correlated. We can use the multivariate normal distribution for this. It behaves similar to a regular normal distribution (which we'd specify a mean and standard deviation). The difference is that in a multivariate normal distribution, there are multiple means that are related to each other through their covariance or correlation matrices. A correlation matrix simply expresses the correlations between the variables - the diagonals are always 1 (because that is the self-correlation) and the off-diagonal cells express the correlation (or covariance depending on the structure) between the 2 variables. Here, we build this by hand by specifying `Rho` (the correlation coefficient). Finally, we use the `rmvnorm` function to generate `nvall` numbers from 2 different normal distributions that have covariance matrix `Sigma` and should have a correlation equal to `Rho`.

We plug these values back into our template using `setValues` and then double-check the results. We expect a little difference between `Rho` and the results of `cor` because of the random number generation, but it is clear that we are pretty close so we can move on.

```{r}

set.seed(42)

# raster template
r_template <- rast(ncol=100, nrow=100, 
                   crs = crs(vect(population_data)), 
                   extent = ext(vect(population_data)))
nvals <- ncell(r_template)

# desired covariance / correlation matrix (2x2)
rho <- 0.7
sd1 <- 1; sd2 <- 2
Sigma <- matrix(c(sd1^2, rho*sd1*sd2, rho*sd1*sd2, sd2^2), nrow=2)

# draw nvals independent bivariate normals
draws <- rmvnorm(n = nvals, sigma = Sigma)  # n x 2 matrix

# convert to rasters
r1 <- setValues(r_template, draws[,1])
r2 <- setValues(r_template, draws[,2])

# quick check empirical correlation (across cells)
cor(values(r1), values(r2))
# plot
plot(c(r1, r2))

rsts <- c(r1,r2)
names(rsts) <- c("var1", "var2")
```


## Changing Aggregation

Now that we have some data with a known relationship, we can start to experiment with the effects of aggregation. To do this efficiently, we'll use a function. I've written out some of the pseudocode below, but I want to point out a few things:

1. In the definition of the function I've specified some defaults (like `pop_cal = "pop"` or `threshold = 1000`). This allows the function to run even if I don't give those values, but they are replaced if I provide any parameters to those arguments.

2. There are two different methods (`greedy` or `random`) for aggregating the data. The greedy algorithm chooses the smallest polygons first, the random chooses polygons at random (subject to seed.)

3. We use the **binary operator** `st_touches` to find all of the polygons touching the polygon indexed by `sf_obj[[idx]]` and the **binary transformer** `st_union` to combine the geometries of the neighboring polygons into a single polygon. We do this in the context of a `group_by`, `summarize` pipeline to combine both the population data and the geometries.

```{r}

aggregate_polygons <- function(sf_obj, pop_col = "pop", threshold = 1000,
                                   method = c("greedy", "random"), seed = NULL) {
  method <- match.arg(method)
  sf_obj <- sf_obj %>%
    mutate(agg_id = row_number(),
           !!pop_col := !!sym(pop_col))
  
  if (!is.null(seed)) set.seed(seed)
  
  repeat {
    below <- which(sf_obj[[pop_col]] < threshold)
    if (length(below) == 0) break
    
    # --- choose starting polygon ---
    if (method == "greedy") {
      idx <- below[which.min(sf_obj[[pop_col]][below])]
    } else if (method == "random") {
      idx <- sample(below, 1)
    }
    
    # find neighbors
    nb <- st_touches(st_make_valid(sf_obj))[[idx]]
    if (length(nb) == 0) {
      warning("Polygon ", idx, " has no neighbors to merge")
      sf_obj[[pop_col]][idx] <- threshold
      next
    }
    
    # --- pick neighbor ---
    if (method == "greedy") {
      nb_idx <- nb[which.min(sf_obj[[pop_col]][nb])]
    } else if (method == "random") {
      nb_idx <- sample(nb, 1)
    }
    
    # merge geometry and population
    sf_obj$geometry[idx] <-st_union(st_make_valid(sf_obj$geometry[c(idx, nb_idx)]))
    sf_obj[[pop_col]][idx] <- sum(sf_obj[[pop_col]][c(idx, nb_idx)])
    sf_obj$agg_id[idx] <- min(sf_obj$agg_id[c(idx, nb_idx)])
    
    # drop merged neighbor
    sf_obj <- sf_obj[-nb_idx, ]
  }
  
  sf_obj %>% 
    group_by(agg_id) %>% 
    summarise(pop = sum(!!sym(pop_col)), geometry = st_union(geometry), .groups = "drop")
}

```

Let's try it out and see how it goes. We'll set a series of thresholds and pass them with `map` to create a list of `sf` objects. We'll then use a second function to generate the maps. Finally, we use `wrap_plots` from `patchwork` to make a multiplanel plot.

```{r}
thresholds <- c(750, 1500, 3000, 6000, 12000)  # population cutoffs
aggregate_pop <- map(thresholds, 
  ~aggregate_polygons(sf_obj = population_data, 
                                    pop_col = "value", threshold = .x,
                                   method = "greedy")
)

plots <- map(aggregate_pop, 
  ~ggplot(data = .x) +
    geom_sf(col = "red") +
    theme_bw()
)

# Combine automatically into a multipanel layout
wrap_plots(plots, ncol = 3)


```


## How Does Aggregation Affect Measured Correlation

```{r}
corr_at_threshold <- function(threshold, polygons, pop_col, r1, method = "greedy", seed = NULL) {
  # aggregate polygons
  agg_polys <- aggregate_polygons(polygons, pop_col, threshold, method = method, seed = seed)
  
  # extract raster values (mean within each polygon)
  v1 <- terra::extract(r1, vect(agg_polys), fun = mean, na.rm = TRUE)[, -1]
  
  tibble(threshold = threshold,
         correlation = cor(v1[,1], v1[,2], use = "complete.obs"))
}
```

```{r}
# --- apply across thresholds ---


results <- map_dfr(thresholds, ~corr_at_threshold(.x, polygons = population_data, "value", r1=rsts, method = "greedy"))

ggplot(results, aes(x = threshold, y = correlation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = rho, linetype = 2, color = "red") +
  theme_minimal() +
  labs(x = "Population aggregation threshold",
       y = "Correlation (r1 vs r2)",
       title = "Effect of polygon aggregation on observed correlation")


```


```{r}


rsts2 <- terra::aggregate(rsts, fact = 4, method = "bilinear")

results2 <- map_dfr(thresholds, ~corr_at_threshold(.x, polygons = population_data, "value", r1=rsts2, method = "greedy"))

ggplot(results2, aes(x = threshold, y = correlation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = rho, linetype = 2, color = "red") +
  theme_minimal() +
  labs(x = "Population aggregation threshold",
       y = "Correlation (r1 vs r2)",
       title = "Effect of polygon aggregation on observed correlation")
```


```{r}

pa_locs <- read_sf("data/original/inclass10/pa.shp") %>% 
  st_transform(., crs=crs(rsts)) %>% 
  st_crop(., population_data)

buffer_dist <- c(10, 1000, 10000, 25000)


pa_buffs <- buffer_dist %>% 
  map(., ~st_buffer(x = pa_locs, dist = .x) %>% 
        mutate(buff_dist = .x) 
        )

plots_buff <- map(pa_buffs, 
  ~ggplot(data = .x) +
    geom_sf(col = "red") +
    theme_bw()
)

# Combine automatically into a multipanel layout
wrap_plots(plots_buff, ncol = 3)



corr_at_buff <- function(polys, r1) {
  # extract raster values (mean within each polygon)
  v1 <- terra::extract(r1, vect(polys), fun = mean, na.rm = TRUE)[, -1]
  
  tibble(buff_dist = unique(polys$buff_dist),
         correlation = cor(v1[,1], v1[,2], use = "complete.obs"))
}

results3 <- map_dfr(pa_buffs, ~corr_at_buff(polys = .x,  r1=rsts))

ggplot(results3, aes(x = buff_dist, y = correlation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = rho, linetype = 2, color = "red") +
  theme_minimal() +
  labs(x = "Population aggregation threshold",
       y = "Correlation (r1 vs r2)",
       title = "Effect of polygon aggregation on observed correlation")
```